---
title: "A note on extraction in the `disordR` package"
author: "Robin Hankin"
date: "2023-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A curious observation

In `disordR_0.0-9-6` and below, the following would be possible:

```
> a <- disord(1:9)
> a[a<3] + a[a>7]
A disord object with hash 141d1b04a56b313bb4a90429ac5b91f69fbd3748 and elements
[1]  9 11
(in some order)
> 
```

It is not immediately clear whether this idiom should return an error
or not.  My first thought was that it should return an error, as
`a[a<3]` has elements 1 and 2 (in some order) and `a[a>7]` has
elements 8 and 9 (in some order) and although it returns 1+8 and 2+9 =
9,11 it could have returned 1+9 and 2+8 = 10,10.  This did not return
an error in versions `<=0.0-9-6`because the hash codes of `a[a<3]` and
`a[a>7]` match.  And it silently returned a result which is, at least
in principle, implementation-specific. At least, I think so.

But reflection it is not clear to me that `a[a<3] + a[a>7]` should
return an error.  It depends on whether you want to think of
"`a[a<3]`" as having an independent existence; specifically, an
existence that is independent of `a[a>7]`. Arguably, both have an
order inherited from `a`. If you consider the order ambiguity of
`a[a<3]` to be inherited purely from `a` then `a[a<3] + a[a>7]` should
not return an error: whatever implementation-specific order a has, the
sum is well-defined up to its own order. If, on the other hand, you
regard the extraction as introducing a new ordering ambiguity, then
`a[a<3] + a[a>7]` is poorly defined [above it was either 9,11 or
10,10] and should return an error.

In versions `>=disordR_0.0-9-6`, idiom `a[a<3] + a[a>7]` does return
an error.  This on the grounds that `disordR` is intended to be used
by packages such as `mvp`, not to be attached.  How does this change
manifest itself in `mvp` idiom?

Under OLD code:

```
> (a <- rmvp(9))
mvp object algebraically equal to
3 a b^2 c^3 d^7 f^6 + a^5 b^5 e^9 f^5 + 8 a^6 b^3 c^11 f^3 + 9 a^6 d^8 f^2 + 7
a^11 b^6 d^2 e^6 f^2 + 4 a^15 e^9 f^6 + 6 b^4 c^5 f^11 + 2 b^5 d^12 f^3 + 5 b^5
e^4 f^6
> coeffs(a)[coeffs(a)<3] + coeffs(a)[coeffs(a)>7]
A disord object with hash d81690c2d9824972ceaa64f5774de95290073820 and elements
[1]  9 11
(in some order)
>
```

It is pretty clear to me that the sum above is not defined. The result
(9 11) is meaningless, or at least as meaningful as 10 10. The
operation corresponding to the idiom does not seem to make sense in
the context of coefficients as managed by the STL map class: "take
every coefficient less than three and add it to every coefficient
greater than 7".  Well yes, but two questions arise:

Firstly, which element of `a[a<3]` is to be matched with which element
of `a[a>7]`?  There are two equally plausible ways of doing it.  The
way that is actually done in the package is to say that both have an
ambiguous but consistent order inherited from `a`.  I don't see any
reason to impose consistency here.

The second question would be to ask what the algebraic meaning of the
sum is.  Something like `coeffs(a)[coeffs(a)<3] <-
coeffs(a)[coeffs(a)<3] + 5` is fine: "take every coefficient less than
3 and add 5 to it".  But although "`a[a<3] + a[a>7]`" might arguably
have a meaning if `a` is an abstract `disord` object, we must ask what
"`coeffs(a)[coeffs(a)<3] + coeffs(a)[coeffs(a)>7]`" means if `a` is an
`mvp` object.  I don't have a good answer to this, at least I cannot
think of an algebraic interpretation in which `9 11` is somehow
"right" and `10 10` is somehow "wrong".



## Note on double square bracket extraction: `x[[i]]`

From version 0.9-2, `disordR` includes `S4` methods for double square
bracket extraction for lists, which simply return an error.  The base
`R` methods for `[[` extract a _single_ element of its argument (as
opposed to '[', which can extract multiple elements).  In many ways
this is inimical to the philosophy of `disordR`, as the elements of a
`disord` list are stored in an implementation-specific order.  Thus

```{r}
(a <- list(1:3,1:9,1:4))
a[[2]]
```

is a perfectly acceptable extraction: we find the second element of
list `a`.  Now consider:


```{r}
library("disordR")
(b <- disord(a))
```

Now we are unable to extract the second element from `b`:


```{r, error=TRUE}
b[[2]]
```

However, there is an `S4` method for  `lapply()`:

```{r}
lapply(b,length)
```

Now, we might ask for the _longest_ element of `a`.  In `R`:


```{r}
l <- unlist(lapply(b,length))  # need to unlist
a[which(l==max(l))]
```





Now this is perfectly well-defined on `disord` objects, but

```{r error=TRUE}
l <- unlist(lapply(b,length))
l
which(l==max(l))
```
